Fisheries example integrating FLR

### 2020-05-26 notebook/cran_vignettes/SI2.Rmd
Recompile the website

---
title: 'Use of the gmse_apply function'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 2)'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
linestretch: 1
---

@@

It is possible to, e.g., specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions (see [SI5](SI5.pdf) for a fisheries example). This can be done by using the default GMSE manager and user functions (written below explicitly, though this is not necessary).

@@

>>>>

---
title: 'Use of the gmse_apply function'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 2)'
author: |
        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
linestretch: 1
---

@@

It is possible to, e.g., specify a simple resource and observation model, but then take advantage of the genetic algorithm to predict policy decisions and user actions (see [Fisheries example integrating FLR](SI5.pdf) for a fisheries example). This can be done by using the default GMSE manager and user functions (written below explicitly, though this is not necessary).

@@

Special considerations for looping with custom sub-models
================================================================================

There are some special considerations that need to be made when using custom sub-models and the `old_list` argument within a loop as above. These considerations boil down to two key points.

1. Custom sub-models *always* need to be read in explicitly as an argument in `gmse_apply` (i.e., they will not be remembered by `old_list`).
2. Custom sub-model arguments also *always* need to be updated *outside* of `gmse_apply` before output is used as an argument in `old_list` (i.e., `gmse_apply` cannot know what custom function argument needs to be updated, so this needs to be done manually).

An example below illustrates the above points more clearly. Assume that the custom resource sub-model defined above needs to be integrated with the default observation, manager, and user sub-models using `gmse_apply`.

The sub-model can be integrated once using `gmse_apply` as demonstrated above, but in the full `gmse_apply` output, the argument `X` will not change from its initial value (because sub-model functions can take any number of arbitrary arguments, `gmse_apply` has no way of knowing that `X` is meant to be the resource number and not some other parameter).

```{r, eval = TRUE}
sim_4 <- gmse_apply(res_mod = alt_res, X = 1000, get_res = "Full");
print(sim_4$basic_output);
```

Note that in the above output, the resource abundance has increased and is now `sim_4$basic_output$resource_results`. But if we look at `sim_4$X`, the value is still `r sim_4$X`.

```{r, eval = TRUE}
print(sim_4$X);
```

To loop through multiple time steps with the custom function `alt_res`, it is therefore necessary to update `sim4$X` with the updated value from either `sim4$resource_vector` or `sim4$basic_output$resource_results` (the two values should be identical). The loop below shows a simple example.

```{r, eval = TRUE}
init_abun   <- 1000;
sim_old     <- gmse_apply(get_res = "Full", res_mod = alt_res, X = init_abun);
for(time_step in 1:20){
    sim_new               <- gmse_apply(res_mod = alt_res, get_res = "Full", 
                                        old_list = sim_old);
    sim_old               <- sim_new;
    sim_old$X             <- sim_new$resource_vector;
}
```

Note again that the custom sub-model is read into to `gmse_apply` as an argument within the loop (`res_mod = alt_res`), and the output of `sim_new` is used to update the custom argument `X` in `alt_res` (`sim_old$X <- sim_new$resource_vector`). The population quickly increases to near carrying capacity, which can be summarised by using the same table structure explained above.

```{r, eval = TRUE}
init_abun   <- 1000;
sim_old     <- gmse_apply(get_res = "Full", res_mod = alt_res, X = init_abun);
sim_sum_4   <- matrix(data = NA, nrow = 5, ncol = 5);
for(time_step in 1:5){
    sim_new                 <- gmse_apply(res_mod = alt_res, get_res = "Full", 
                                          old_list = sim_old);
    sim_sum_4[time_step, 1] <- time_step;
    sim_sum_4[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_4[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_4[time_step, 4] <- sim_new$basic_output$manager_results[3];
    sim_sum_4[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
    sim_old                 <- sim_new;
    sim_old$X               <- sim_new$resource_vector;
}
colnames(sim_sum_4) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", 
                         "Cull_count");
print(sim_sum_4);
```

This is the recommended way to loop custom functions in `gmse_apply`. Note that elements of `old_list` will over-ride custom arguments to `gmse_apply` so **specifying custom arguments that are already present in `old_list` will not work**.

### 2020-05-26 notebook/cran_vignettes/SI4.Rmd
Recompile the website

---
title: 'Advanced case study options'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 4)'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, 
  O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  html_document: default
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: 
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
biblio-style: apalike
---

@@

Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`. To do this, we use the same scenario described in [SI3](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement). We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.

@@

We now recreate the simulations in [SI3](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`. Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`). Following the protocol introduced in [SI2](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.

@@

We can then loop over 30 time steps to recreate the simulations from [SI3](SI3.pdf). In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum_1`. The inclusion of the argument `old_list` tells `gmse_apply` to use parameters and values from the list `sim_old` in the new time step.

@@

The above output from `sim_sum_1` shows the data frame that holds the information we were interested in pulling out of our simulation results. All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record. It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user sub-models. Custom sub-models could produce different outputs in `sim_new` (see [SI2](SI2.pdf) for examples). For default sub-models, there are some list elements that might be especially useful. These elements can potentially be edited *within the above loop* to dynamically adjust simulations. For more explanation of built-in GMSE data arrays, see [SI7](SI7.pdf).

- [`sim_new$resource_array`](SI7.pdf#resources): A table holding all information on resources. Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?), (21) how much yield has the resource consumed, and (22) how many times the resource can consume yield in one time step.
- [`sim_new$AGENTS`](SI7.pdf#agents): A table holding basic information on agents (manager and users). Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) baseline budget (`manager_budget` and `user_budget`), (18-24) agent's perception of the efficacy of scaring, culling, castrating, feeding, helping, tending crops, and killing crops, (25-26) increments to budget, (27) unused.
- `sim_new$observation_vector`: Estimate of total resource number from the observation model ([`observation_array`](SI7.pdf#observation) also holds this information in a different way depending on `observe_type`)
- [`sim_new$LAND`](SI7.pdf#land): The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers. Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default sub-models, but could be used to store values that affect resources and agents. Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for the actions of each agent can be set for each user in [`sim_new$manager_array`](SI7.pdf#cost).
- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in [`sim_new$user_array`](SI7.pdf#action).

@@

```{r, fig.cap = "Default position of land ownership by farmers."}
image(x = sim_old$LAND[,,3], col = topo.colors(9), xaxt = "n", yaxt = "n");
```

@@

In this example, we showed how the built-in resource, observation, manager, and user sub-models can be customised by manipulating the data within the data structures that they use. The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations. Readers seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to [SI2](SI2.pdf) that introduces `gmse_apply` more generally. Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

>>>>

---
title: 'Advanced case study options'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 4)'
author: |
        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: 
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
biblio-style: apalike
---

@@

Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`. To do this, we use the same scenario described in [Example case study in GMSE](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement). We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.

@@

We now recreate the simulations in [Example case study in GMSE](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`. Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`). Following the protocol introduced in [Use of the gmse_apply function](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.

@@

We can then loop over 30 time steps to recreate the simulations from [Example case study in GMSE](SI3.pdf). In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum_1`. The inclusion of the argument `old_list` tells `gmse_apply` to use parameters and values from the list `sim_old` in the new time step.

@@

The above output from `sim_sum_1` shows the data frame that holds the information we were interested in pulling out of our simulation results. All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record. It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user sub-models. Custom sub-models could produce different outputs in `sim_new` (see [Use of the gmse_apply function](SI2.pdf) for examples). For default sub-models, there are some list elements that might be especially useful. These elements can potentially be edited *within the above loop* to dynamically adjust simulations. For more explanation of built-in GMSE data arrays, see [Default GMSE data structures](SI7.pdf).

- [`sim_new$resource_array`](SI7.pdf): A table holding all information on resources. Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?), (21) how much yield has the resource consumed, and (22) how many times the resource can consume yield in one time step.
- [`sim_new$AGENTS`](SI7.pdf): A table holding basic information on agents (manager and users). Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) baseline budget (`manager_budget` and `user_budget`), (18-24) agent's perception of the efficacy of scaring, culling, castrating, feeding, helping, tending crops, and killing crops, (25-26) increments to budget, (27) unused.
- `sim_new$observation_vector`: Estimate of total resource number from the observation model ([`observation_array`](SI7.pdf) also holds this information in a different way depending on `observe_type`)
- [`sim_new$LAND`](SI7.pdf): The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers. Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default sub-models, but could be used to store values that affect resources and agents. Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for the actions of each agent can be set for each user in [`sim_new$manager_array`](SI7.pdf).
- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in [`sim_new$user_array`](SI7.pdf).

@@

```{r, fig.cap = "Default position of land ownership by farmers.", fig.height = 6, fig.width = 6}
image(x = sim_old$LAND[,,3], col = topo.colors(9), xaxt = "n", yaxt = "n");
```

@@

In this example, we showed how the built-in resource, observation, manager, and user sub-models can be customised by manipulating the data within the data structures that they use. The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations. Readers seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to [Use of the gmse_apply function](SI2.pdf) that introduces `gmse_apply` more generally. Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

### 2020-05-26 notebook/cran_vignettes/SI5.Rmd
Recompile the website

```{r, fig.height = 5.2, fig.cap = "Output of the FLR plot function for an example Ricker model of stock recruitment on North Sea Herring data."}
plot(newFL, cex = 0.7);
```

>>>>

```{r, fig.height = 5.2, fig.cap = "Output of the FLR plot function for an example Ricker model of stock recruitment on North Sea Herring data.", message = FALSE}
plot(newFL, cex = 0.7);
```

### 2020-05-26 notebook/ms/SI5.Rmd
Recompile the website

As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of animal populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can be managed and mitigated. Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm](SI1.pdf). Such integration might be possible using the `gmse_apply` function, which allows for [custom defined sub-models](SI2.pdf) to be used within the GMSE framework, and with default GMSE sub-models. Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing competing stakeholder demands and conflicts. We do not attempt such an ambitious project here, but instead show how such a project could be developed through integration of [FLR](http://www.flr-project.org/) and `gmse_apply`.

@@

The FLR toolset includes [a series of packages](http://www.flr-project.org/#packages), with several [tutorials](http://www.flr-project.org/doc/index.html) for using them. For simplicity, we focus on [a model of stock recruitment](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) to be used as the population model in `gmse_apply`. This population model will use sample data and one of the many available stock-recruitment models available in FLR, and a custom function will be written to return a single value for stock recruitment. Currently, `gmse_apply` requires that sub-models return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.pdf) as GMSE sub-models. But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch). For simplicity, the observation (i.e., estimation) model will be the stock reported from the population model with error. The manager and user models, however, will employ the full power of the default GMSE functions to simulate management and user actions. We first show how a custom function can be made that applies the FLR toolset to a population model.

>>>>

As part of the [ConFooBio](https://sti-cs.org/confoobio/) project, a central focus of GMSE is on simulating the management of animal populations of conservation interest, with a particular emphasis on understanding conservation conflict; further development of GMSE is expected to continue with this as a priority, further building upon the decision-making algorithms of managers and users to better understand how conflict arises and can be managed and mitigated. Hence, GMSE is not intended as a substitute for packages such as [FLR](http://www.flr-project.org/), but the integration of these packages with GMSE could make use of GSME's current and future simulation capabilities, and particularly the [genetic algorithm](SI1.html). Such integration might be possible using the `gmse_apply` function, which allows for [custom defined sub-models](SI2.html) to be used within the GMSE framework, and with default GMSE sub-models. Hence, GMSE might be especially useful for modelling the management of fisheries under conditions of increasing competing stakeholder demands and conflicts. We do not attempt such an ambitious project here, but instead show how such a project could be developed through integration of [FLR](http://www.flr-project.org/) and `gmse_apply`.

@@

The FLR toolset includes [a series of packages](http://www.flr-project.org/#packages), with several [tutorials](http://www.flr-project.org/doc/index.html) for using them. For simplicity, we focus on [a model of stock recruitment](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) to be used as the population model in `gmse_apply`. This population model will use sample data and one of the many available stock-recruitment models available in FLR, and a custom function will be written to return a single value for stock recruitment. Currently, `gmse_apply` requires that sub-models return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.html) as GMSE sub-models. But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch). For simplicity, the observation (i.e., estimation) model will be the stock reported from the population model with error. The manager and user models, however, will employ the full power of the default GMSE functions to simulate management and user actions. We first show how a custom function can be made that applies the FLR toolset to a population model.
